Discrete Probability for Discrete Quantum Theories

### 2016-05-06 prob.tex
mess

\makeatletter
\newtheoremstyle{indented}
  {3pt}% space before
  {3pt}% space after
  {\addtolength{\@totalleftmargin}{3.5em}
   \addtolength{\linewidth}{-3.5em}
   \parshape 1 3.5em \linewidth}% body font
  {}% indent
  {\bfseries}% header font
  {.}% punctuation
  {.5em}% after theorem header
  {}% header specification (empty for default)
\makeatother
\theoremstyle{indented}
%% \theoremstyle{remark}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{cor}{Corollary}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\qevents}{\ensuremath{\mathcal{E}}}
\newcommand{\pmeas}{\ensuremath{\mu}}
\newcommand{\Hilb}{\mathcal{H}}
\newcommand{\pr}[2]{\langle #1, #2 \rangle}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ip}[2]{\langle #1 | #2 \rangle}
\newcommand{\proj}[1]{|#1 \rangle\langle #1 |}
\newcommand{\ps}{\texttt{+}}
\newcommand{\ms}{\texttt{-}}
\newcommand{\poss}{{\mbox{\wesa{possible}}}}
\newcommand{\imposs}{{\mbox{\wesa{impossible}}}}
\newcommand{\likely}{{\mbox{\wesa{likely}}}}
\newcommand{\unlikely}{{\mbox{\wesa{unlikely}}}}
\newcommand{\necess}{{\mbox{\wesa{necessary}}}}
\newcommand{\overflow}{{\mbox{\wesa{overflow}}}}

@@

\begin{document}
\title{Discrete Probability for Discrete Quantum Computing}
\author{}
\date{\today}
\maketitle

@@

We review the conventional presentation of probability spaces and then
discuss several variations that avoid using the real interval $[0,1]$.

@@

A \emph{probability
  space}~\cite{inun.425605319950101,GrahamKnuthPatashnik1994,rohatgi2011introduction}
specifies the necessary conditions for reasoning coherently about
collections of uncertain events. It consists of a \emph{sample space}
$\Omega$, a space of \emph{events}~$\events$, and a \emph{probability
  measure}~$\pmeas$. In this paper, we will only consider
\emph{finite} sets of events: we therefore define a sample space
$\Omega$ as an arbitrary non-empty finite set and the space of events
$\events$ as $2^\Omega$, the powerset of $\Omega$. Given the set of
events $\events$, a \emph{probability measure} is a function
$\pmeas : \events \rightarrow [0,1]$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = 1$, and 
\item for a collection $E_i$ of pairwise disjoint events, 
  $\pmeas(\bigcup_i E_i) = \sum_i \pmeas(E_i)$.
\end{itemize}

@@

In a strict computational or experimental setting, one may question
the reliance of the definition of probability space on the uncountable
and uncomputable real interval $[0,1]$. This interval includes numbers
like $0.h_1h_2h_3\ldots$ where $h_i$ is 1 or 0 depending on whether
Turing machine $M_i$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases.
% \yutsung{Check the meaning of the ``computing $\frac{\pi}{4}$'', because
% Bailey Borwein Plouffe formula can computing the $n$th binary digit of $\pi$ 
% using base 16 directly.}
Therefore, in a resource-aware setting, it is more appropriate to
consider probability measures that map events to a finite set of
elements computable with a fixed set of resources. We will consider
two approaches: set-valued probability
measures~\cite{Artstein1972,PuriRalescu1983} and interval-valued
probability
measures~\cite{Dempster1967,Zadeh1986,Weichselberger2000,JamisonLodwick2004}.

@@

Instead of using every point in the real interval $[0,1]$ we can
consider subsets~$\mathscr{L}$ of $[0,1]$ and only consider probability
measures up to set membership with intuitive addition\footnote{Notice that the set-valued measure in~\cite{Artstein1972,PuriRalescu1983},
only require $I_{1}+I_{2}=\left\{ a_{1}+a_{2}\middle|a_{1}\in I_{1},a_{2}\in I_{2}\right\} $
because they focused on measure theory. In probability thoery, $2\in\pmeas\left(E\right)$
doesn't really make sense so that we consider intersect $[0,1]$ in
our definition.} 
\[
I_{1}+I_{2}=\left\{ a_{1}+a_{2}\middle|a_{1}\in I_{1},a_{2}\in I_{2}\right\} \cap[0,1]\textrm{ .}
\]
Then, a probability measure in this case is modified to a function
$\pmeas:\events\rightarrow\mathscr{L}$ such that: 
\begin{itemize}
\item $1\in\pmeas(\Omega)$, and 
\item for a collection $E_{i}$ of pairwise disjoint events, $\pmeas(\bigcup_{i}E_{i})=\sum_{i}\pmeas(E_{i})$. 
\end{itemize}
The simplest such situation is to partition the interval $[0,1]$
into $\{0\}$ (which we will call $\imposs$) and the half-open interval
$(0,1]$ (which we will call $\poss$). We will call the resulting
set $\left\{ \imposs,\poss\right\} $, the set $\mathscr{L}_{2}$.

@@

Thus if we have an event $E$ with probability $[a,b]$ where $a=0.1$
and $b=0.7$, we have that: 
\begin{itemize}
\item the strength of evidence supporting $E$ is 0.1; since either $E$
or its complement must happen, we conclude that there is 0.9 evidence
supporting the complement of $E$; 
\item the strength of evidence contradicting $E$ is 0.7; again since either
$E$ or its complement must happen, we conclude that there is 0.3
evidence contradicting the complement of $E$. 
\end{itemize}
\yutsung{Do we use the law of excluded middle here? You remind me
Agda : ) Did Homotopy Type Theory people said anything about the probability?}\\
Turning things around, the strength of evidence that contradicts $E$
is evidence supporting the complement of $E$. The complement of $E$
must therefore have probability $\left[1-b,1-a\right]$ which we abbreviate
$1-\left[a,b\right]$, so the probability measure $\pmeas$ should
satisfy:
\begin{itemize}
\item $\pmeas(\emptyset)=[0,0]$, 
\item $\pmeas(\Omega)=[1,1]$, and 
\item $\pmeas\left(\Omega\backslash E\right)=1-\pmeas\left(E\right)$
\end{itemize}
However, if we want $\pmeas(\Omega)=[1,1]$, and the probability assignment
for singlton sets $\pmeas(\{HH\})=\unlikely$, $\pmeas(\{HT\})=\imposs$,
$\pmeas(\{TH\})=\likely$, and $\pmeas(\{TT\})=\imposs$ as in example~\ref{Two-coin-set-valued-again},
we only have $\pmeas(\Omega)\subsetneq\pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})$.
In general, we only require
\begin{itemize}
\item for a collection $E_{i}$ of pairwise disjoint events, we have $\pmeas\left(\bigcup_{i}E_{i}\right)\subseteq\sum_{i}\pmeas\left(E_{i}\right)$. 
\end{itemize}
This statement means that the evidences of $\bigcup_{i}E_{i}$ is
at least as strong as putting all the evidences of $E_{i}$ together,
but some evidence may only be acquired for $\bigcup_{i}E_{i}$ as
the whole. Therefore, $\pmeas\left(\bigcup_{i}E_{i}\right)$ is a
subset of $\sum_{i}\pmeas\left(E_{i}\right)$, but may not equal.
In our example, 
\begin{eqnarray*}
 &  & \pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})\\
 & = & \imposs+\unlikely+\imposs+\likely\\
 & = & \left[0,0\right]+\left[0,\frac{1}{2}\right]+\left[0,0\right]+\left[\frac{1}{2},1\right]=\left[\frac{1}{2},\frac{3}{2}\right]
\end{eqnarray*}
The above equation told us $\pmeas(\Omega)\subseteq\left[\frac{1}{2},\frac{3}{2}\right]$.
However, because of $\pmeas(\emptyset)=\imposs$, we have $\pmeas\left(\Omega\right)=1-\pmeas\left(\emptyset\right)=\necess$,
but $\pmeas(\emptyset)=\imposs$ cannot be used to reasoning the probability
of $\{HH\}$, $\{HT\}$, $\{TH\}$ and $\{TT\}$ individually. The
full probability assignment is shown in the following example.

@@

% Now consider the event
% that some employee's age is in the range $\{23,24\}$. Reasoning as
% above, the probability of this event is $[\frac{1}{3},\frac{1}{3}]$ as
% only the first employee qualifies. Clearly if we were to ask the
% probability that some employee's age is in the range
% $\{20,21,22,23,24\}$ the result should be $[1,1]$ as this range covers
% all the possibilities. The way to calculate this result from the two
% previous ones is as follows: the evidence for the combined event to be
% necessary is at least as strong as the evidence that each disjoint
% event that contributes to it is necessary. So the lower bound
% probability for the combined event is at least
% $\frac{1}{3}$. Similarly the upper bound probability is at least
% $1$. But now we can reason using the complements of the events about
% the necessity and possibility of refuting each event. For the event
% that some employee's age is \emph{not} in the range $\{20,21,22\}$ we
% have a probability $[\frac{1}{3},\frac{1}{3}]$ because it is necessary
% that $A_1$ is not in that range. Similary, the event that some employee's age is
% \emph{not} in the range $\{23,24\}$ is
% $[\frac{2}{3},\frac{2}{3}]$. Thus although the positive evidence for
% necessity

>>>>

% \makeatletter
% \newtheoremstyle{indented}
%   {3pt}% space before
%   {3pt}% space after
%   {\addtolength{\@totalleftmargin}{3.5em}
%    \addtolength{\linewidth}{-3.5em}
%    \parshape 1 3.5em \linewidth}% body font
%   {}% indent
%   {\bfseries}% header font
%   {.}% punctuation
%   {.5em}% after theorem header
%   {}% header specification (empty for default)
% \makeatother
% \theoremstyle{indented}
\theoremstyle{remark}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{cor}{Corollary}
\newcommand{\events}{\ensuremath{\mathcal{E}}}
\newcommand{\qevents}{\ensuremath{\mathcal{E}}}
\newcommand{\pmeas}{\ensuremath{\mu}}
\newcommand{\Hilb}{\mathcal{H}}
\newcommand{\pr}[2]{\langle #1, #2 \rangle}
\newcommand{\ket}[1]{|#1\rangle}
\newcommand{\bra}[1]{\langle#1|}
\newcommand{\ip}[2]{\langle #1 | #2 \rangle}
\newcommand{\proj}[1]{|#1 \rangle\langle #1 |}
\newcommand{\ps}{\texttt{+}}
\newcommand{\ms}{\texttt{-}}
\newcommand{\poss}{{\mbox{\wesa{possible}}}}
\newcommand{\imposs}{{\mbox{\wesa{impossible}}}}
\newcommand{\likely}{{\mbox{\wesa{likely}}}}
\newcommand{\unlikely}{{\mbox{\wesa{unlikely}}}}
\newcommand{\necess}{{\mbox{\wesa{necessary}}}}
\newcommand{\overflow}{{\mbox{\wesa{overflow}}}}

@@

\begin{document}
\title{Discrete Probability for Discrete Quantum Theories}
\author{}
\date{\today}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

A \emph{probability space} specifies the necessary conditions for
reasoning coherently about collections of uncertain events.  We review
the conventional presentation of probability spaces and then discuss
several variations that avoid using the real interval $[0,1]$.

@@

The conventional definition of a probability
space~\cite{inun.425605319950101,GrahamKnuthPatashnik1994,rohatgi2011introduction}
builds upon the real numbers. In more detail, a probability space
consists of a \emph{sample space} $\Omega$, a space of
\emph{events}~$\events$, and a \emph{probability measure}~$\pmeas$
mapping events in $\events$ to the real interval $[0,1]$. In this
paper, we will only consider \emph{finite} sets of events: we
therefore restrict our attention to non-empty finite sets $\Omega$ as
the sample space. The space of events $\events$ includes every
possible subset of $\Omega$: it is the powerset~$2^\Omega$. Given the
set of events $\events$, a \emph{probability measure} is a function
$\pmeas : \events \rightarrow [0,1]$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = 1$, and 
\item for a collection $E_i$ of pairwise disjoint events, 
  $\pmeas(\bigcup_i E_i) = \sum_i \pmeas(E_i)$.
\end{itemize}

@@

In a strict computational or experimental setting, one may question
the reliance of the definition of probability space on the uncountable
and uncomputable real interval $[0,1]$. This interval includes numbers
like $0.h_1h_2h_3\ldots$ where $h_i$ is 1 or 0 depending on whether
Turing machine $M_i$ halts or not. Such numbers cannot be
computed. This interval also includes numbers like $\frac{\pi}{4}$
which can only be computed with increasingly large resources as the
precision increases.
% \yutsung{Check the meaning of the ``computing $\frac{\pi}{4}$'', because
% Bailey Borwein Plouffe formula can computing the $n$th binary digit of $\pi$ 
% using base 16 directly.}
Therefore, in a resource-aware computational or experimental setting,
it is more appropriate to consider probability measures that map
events to a finite set of elements computable with a fixed set of
resources. We will consider two approaches from the literature:
set-valued probability measures~\cite{Artstein1972,PuriRalescu1983}
and interval-valued probability
measures~\cite{Dempster1967,Zadeh1986,Weichselberger2000,JamisonLodwick2004}.

@@

Instead of using every point in the real interval $[0,1]$ as a
potential value for a probability measure, we can lump points together
in sets and consider probability measures up to set membership. The
simplest such situation is to partition the interval $[0,1]$ into two
sets: the set $\{0\}$ (which we will call $\imposs$) and the set
containing all the points in the half-open interval $(0,1]$ (which we
will call $\poss$). We will call the resulting collection
$\left\{ \imposs,\poss\right\} $, the collection $\mathscr{L}_{2}$. To
completely specify a probability space, we also need to specify how to
``add'' the probabilities in $\mathscr{L}_2$. Given two subsets of the
real interval, $I_1$ and $I_2$, the general formula for adding them
is:\footnote{Notice that the set-valued measure
  in~\cite{Artstein1972,PuriRalescu1983}, only require
  $I_{1}+I_{2}=\left\{ a_{1}+a_{2}\middle|a_{1}\in I_{1},a_{2}\in
    I_{2}\right\} $
  because they focused on measure theory. In probability theory,
  $2\in\pmeas\left(E\right)$ doesn't really make sense so that we
  consider intersect $[0,1]$ in our definition.}
\[
I_{1} \vee I_{2}=\left\{ x+y~\middle|~ x \in I_{1}, y \in I_{2}\right\} ~\cap~ [0,1]\textrm{ .}
\]
In the case of $\mathscr{L}_2$, this reduces to $x \vee y = \imposs$
iff $x = y = \imposs$.

\amr{
From Puri and Ralescu's paper: Fix $n=1$ (only 1-dimension). Conditions on $\mu$:
\begin{itemize}
\item $\mu(E) \neq \emptyset$ for every $E \in \events$, 
\item $1 \in \mu(\Omega)$, 
\item for every disjoint family $A_i$ we have, $\mu(\bigcup_i A_i) =
  \sum_i \mu(A_i)$
\end{itemize}
where the sum of subsets of $\mathbb{R}$ (not $\mathbb{R}^n$ as in the
paper) is as follows. Let $X$ and $Y$ be subsets of $\mathbb{R}$, then
$X + Y = \{ x + y ~|~ x \in X, y \in Y \}$
}

To summarize, an $\mathscr{L}$-valued probability is a function
$\pmeas:\events\rightarrow\mathscr{L}$ such that:
\begin{itemize}
\item $1\in\pmeas(\Omega)$, and 
\item for a collection $E_{i}$ of pairwise disjoint events, $\pmeas(\bigcup_{i}E_{i})=\bigvee_i\pmeas(E_{i})$. 
\end{itemize}

@@

Thus if we have an event $E$ with probability $[a,b]$ where $a=0.1$
and $b=0.7$, we have that: 
\begin{itemize}
\item the strength of evidence supporting $E$ is 0.1; since either $E$
or its complement must happen, we conclude that there is 0.9 evidence
supporting the complement of $E$; 
\item the strength of evidence contradicting $E$ is 0.7; again since either
$E$ or its complement must happen, we conclude that there is 0.3
evidence contradicting the complement of $E$. 
\end{itemize}
\yutsung{Do we use the law of excluded middle here? You remind me
Agda : ) Did Homotopy Type Theory people said anything about the probability?}\\
Turning things around, the strength of evidence that contradicts $E$
is evidence supporting the complement of $E$. The complement of $E$
must therefore have probability $\left[1-b,1-a\right]$ which we abbreviate
$1-\left[a,b\right]$, so the probability measure $\pmeas$ should
satisfy:
\begin{itemize}
\item $\pmeas(\emptyset)=[0,0]$, 
\item $\pmeas(\Omega)=[1,1]$, and 
\item $\pmeas\left(\Omega\backslash E\right)=1-\pmeas\left(E\right)$
\end{itemize}
However, if we want $\pmeas(\Omega)=[1,1]$, and the probability assignment
for singleton sets $\pmeas(\{HH\})=\unlikely$, $\pmeas(\{HT\})=\imposs$,
$\pmeas(\{TH\})=\likely$, and $\pmeas(\{TT\})=\imposs$ as in example~\ref{Two-coin-set-valued-again},
we only have $\pmeas(\Omega)\subsetneq\pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})$.
In general, we only require
\begin{itemize}
\item for a collection $E_{i}$ of pairwise disjoint events, we have $\pmeas\left(\bigcup_{i}E_{i}\right)\subseteq\sum_{i}\pmeas\left(E_{i}\right)$. 
\end{itemize}
This statement means that the evidences of $\bigcup_{i}E_{i}$ is
at least as strong as putting all the evidences of $E_{i}$ together,
but some evidence may only be acquired for $\bigcup_{i}E_{i}$ as
the whole. Therefore, $\pmeas\left(\bigcup_{i}E_{i}\right)$ is a
subset of $\sum_{i}\pmeas\left(E_{i}\right)$, but may not equal.
In our example, 
\begin{eqnarray*}
 &  & \pmeas(\{HH\})+\pmeas(\{HT\})+\pmeas(\{TH\})+\pmeas(\{TT\})\\
 & = & \imposs+\unlikely+\imposs+\likely\\
 & = & \left[0,0\right]+\left[0,\frac{1}{2}\right]+\left[0,0\right]+\left[\frac{1}{2},1\right]=\left[\frac{1}{2},\frac{3}{2}\right]
\end{eqnarray*}
The above equation told us $\pmeas(\Omega)\subseteq\left[\frac{1}{2},\frac{3}{2}\right]$.
However, because of $\pmeas(\emptyset)=\imposs$, we have $\pmeas\left(\Omega\right)=1-\pmeas\left(\emptyset\right)=\necess$,
but $\pmeas(\emptyset)=\imposs$ cannot be used to reasoning the probability
of $\{HH\}$, $\{HT\}$, $\{TH\}$ and $\{TT\}$ individually. The
full probability assignment is shown in the following example.

@@

% Now consider the event
% that some employee's age is in the range $\{23,24\}$. Reasoning as
% above, the probability of this event is $[\frac{1}{3},\frac{1}{3}]$ as
% only the first employee qualifies. Clearly if we were to ask the
% probability that some employee's age is in the range
% $\{20,21,22,23,24\}$ the result should be $[1,1]$ as this range covers
% all the possibilities. The way to calculate this result from the two
% previous ones is as follows: the evidence for the combined event to be
% necessary is at least as strong as the evidence that each disjoint
% event that contributes to it is necessary. So the lower bound
% probability for the combined event is at least
% $\frac{1}{3}$. Similarly the upper bound probability is at least
% $1$. But now we can reason using the complements of the events about
% the necessity and possibility of refuting each event. For the event
% that some employee's age is \emph{not} in the range $\{20,21,22\}$ we
% have a probability $[\frac{1}{3},\frac{1}{3}]$ because it is necessary
% that $A_1$ is not in that range. Similarly, the event that some employee's age is
% \emph{not} in the range $\{23,24\}$ is
% $[\frac{2}{3},\frac{2}{3}]$. Thus although the positive evidence for
% necessity
